Solution-aware vs global ReLU selection: \\ partial MILP strikes back for DNN verification

### 2025-04-19 ATVA 2025/Comparison.tex
comparison

\noindent {\em Running Example:} Consider Fig. \ref{img:FSB_example}. 
It has no bias, so $s_{SR}=s_{FSB}$.
We have $\lambda(b)=d$, $\lambda(a)=\frac{cd}{2}$.
Consider the real improvement on the {\em maximal bound} computed by pMILP$_X(z)$ when opening the $\ReLU$ of node $a$ ($X=\{a\}$) compared with $X=\emptyset$, and compare with 
$s_{FSB}(a)$. Let us do a case analysis:

The most favorable cases are $c < 0 < d$ and $d < 0 < c$: 
as $cd <0$, we have $s_{FSB}(a)=\max(0,\frac{cd}{4}) = 0$.
Because $cd <0$, both $a$ and $\hat{a}$ need to be minimized by pMILP in order to maximize the value of $z$. The LP approximation ($X=\emptyset$) matches the triangle approximation represented in Fig. \ref{triangle}: having a value $sol(a)$ for $a$, the minimal value for $\hat{a}$ is $\ReLU(sol(a))$. If we open $a$ ($X=\{a\}$), then the value for $\hat{a}$ will also be $\ReLU(sol(a))$. That is, opening this ReLU is not helping in computing $\UB(z)$, as correctly predicted by the score $s_{FSB}(a)= 0$.

@@

	\end{tikzpicture}
		\caption{Triangle approximation
		\vspace{-0.3cm}}
	\label{triangle}
	\end{centering}
	\end{figure}

Case $c>0,d>0$: $s_{FSB}(a)=\frac{cd}{4}$.
The value of $a,\hat{a},b,\hat{b}$ should be maximized to maximize the value of $z$, because $W_{bz}=d>0$ and $W_{ab}=c>0$. 
%Each variation of $b$ by $\Delta(b)$ will have an impact $r(b) \cdot \Delta(b)=\frac{\Delta(b)}{2}$ on the value of $\hat{b}$, and thus $\frac{d \Delta(b)}{2}$ on $z$. 
Now, let us call $sol(a)$ the maximum value for $a$.
The maximum value for $\hat{a}$ is $r(a)\cdot (sol(a)-LB(a))$ according to the triangle approximation when $\ReLU(a)$ is not opened. Now, if $\ReLU(a)$ is opened, then 
the value of $\hat{a}$ is $\ReLU(sol(a))$: the difference is at most $r(a) \cdot LB(a) = \frac{1}{2}$. The difference $\Delta(b)$ for the value of $b$ is thus at most $\frac{c}{2}$, which means a difference at most $\frac{c}{4}$ for $\hat{b}$, using the upper function of the triangle approximation of rate $r(b)$, as the value of $\hat{b}$ should be maximized.
This means a difference of at most $\frac{cd}{4}$ on the value of $z$, which is what is computed by $s_{FSB}(a)$. Of course, the difference of opening $\ReLU(a)$ could also be 0 e.g. if $\sol(a)=\LB(a)$ or $\sol(a)=\UB(a)$.

The last case is the most problematic: 
Case $c<0,d<0$, implying that $s_{FSB}(a)=\frac{cd}{4}$, because $cd >0$.
As for the case of $c>0, d>0$, opening $\ReLU(a)$ will have an impact on the value of $\hat{a}$,
at most $\frac{1}{2}$, and hence a change $\Delta(b)$ of at most $\frac{c}{2}$ on the value of $b$.
However, as $d<0$, value of $\hat{b}$ needs to be minimized. That is, the value of $\hat{b}$ will be the value of $\ReLU(sol(b))$, and the change in $\hat{b}$ will be either 0 in case $sol(b)<0$ (phase 0 of ReLU),
or $\Delta(b)<\frac{c}{2}$ otherwise (phase Id of ReLU), that is, $z$ will be modified by 
either 0 or $d \Delta(b)<\frac{cd}{2}$, as compared with $s_{FSB}(a)=\frac{cd}{4}$.

We call {\em global scoring} (GS) such functions $s_{FSB},s_{SR}$  because they score ReLUs as accuratly as possible, considering that they do not have access to the solution values $sol(a),sol(b)$. This analysis already hints at our novel ranking function, that will be solution-aware to be more accurate.

>>>>

	{\color{red}
\noindent {\em Running Example:} Consider Fig. \ref{img:FSB_example}. 
It has no bias, so $s_{SR}=s_{FSB}$. We have $\lambda(b)=d$, $\lambda(a)=\frac{cd}{2}$.
The value of $s_{FSB}(a)$ depends on the signs of $c,d$.

Compare $s_{FSB}(a)$ with $imp(a)$, the real improvement on the {\em maximal bound} computed by pMILP$_X(z)$ when opening the $\ReLU$ of node $a$ ($X=\{a\}$), yielding 
$\val(\hat{a}) = \ReLU(\val(a))$, versus having $X=\emptyset$, for which 
$\val(\hat{a})$ can be any value in the triangle approximation (Figure \ref{triangle})), 
according to Prop. \ref{lp}.

The most favorable cases are $c < 0 < d$ and $d < 0 < c$: 
as $cd <0$, we have $s_{FSB}(a)=\max(0,\frac{cd}{4}) = 0$.
Because $cd <0$, both $a$ and $\hat{a}$ need to be minimized by MILP$_X$ in order to maximize the value of $z$. For $X=\emptyset$, the LP approximation (=MILP$_\emptyset$) will thus 
set $\val(\hat{a}) = \ReLU(\val(a))$ as this is the minimum value in the triangle approximation. Notice that opening $X=\{a\}$ yields the same $\val(\hat{a}) = \ReLU(\val(a))$.
That is, opening node $a$ is not improving the bound $\UB(z)$, as correctly predicted by the score $s_{FSB}(a)= imp(a) = 0$.

@@

	\end{tikzpicture}
		\caption{Triangle approximation with $r(a)=\frac{1}{2}$
		\vspace{-0.3cm}}
	\label{triangle}
	\end{centering}
	\end{figure}

Case $c>0,d>0$: we have $s_{FSB}(a)=\frac{cd}{4}$.
The value of $a,\hat{a},b,\hat{b}$ should be maximized to maximize the value of $z$, because $W_{bz}=d>0$ and $W_{ab}=c>0$. 
%Each variation of $b$ by $\Delta(b)$ will have an impact $r(b) \cdot \Delta(b)=\frac{\Delta(b)}{2}$ on the value of $\hat{b}$, and thus $\frac{d \Delta(b)}{2}$ on $z$. 
Now, let us call $val(a)$ the maximum value for $a$ 
(same for MILP$_\emptyset$ and MILP$_{\{a\}}$).
The maximum value for $\hat{a}$ under LP ($X=\emptyset$) is 
$\frac{1}{2} \cdot (\val(a)-LB(a))$ following the triangle approximation. 
Now, if $X=\{a\}$, then the value of $\hat{a}$ is $\ReLU(\val(a))$: the difference is 
between 0 (for $\val(a)\in \{LB,UB\}$) and $\frac{1}{2}$ (for $\val(a)=0$). 
The difference $\Delta(b)$ for the value of $b$ is thus between 0 and $\frac{c}{2}$, which means a difference at most $\frac{c}{4}$ for $\hat{b}$, using the upper function of the triangle approximation of rate $r(b)$, as the value of $\hat{b}$ should be maximized.
This means a difference $imp(a) \in [0, \frac{cd}{4}]$ on the value of $z$ (depending on the maximal value $val(a)$), to compare with the fixed $s_{FSB}(a)=\frac{cd}{4}$.

The last case is the most problematic: 
Case $c<0,d<0$, implying that $s_{FSB}(a)=\frac{cd}{4}$, because $cd >0$.
Opening $\ReLU(a)$ will have the same impact on the value of $\hat{a}$ then $b$
than in case $c>0,d>0$, with difference of $\val(b)$ between $[0,\frac{c}{2}]$.
However, as $d<0$, value of $\hat{b}$ needs to be minimized to maximize the value of $z$. That is, the value of $\hat{b}$ will be the value of $\ReLU(sol(b))$, and the change in $\hat{b}$ will be either 0 in case $\val(b)<0$,
or $\Delta(b) \in [0,\frac{c}{2}]$ for $\val(b)>0$.
That is, $z$ will be modified by either $imp(a)=0$ or $imp(a) = d \Delta(b)$, to be compared with the fix value $s_{FSB}(a)=\frac{cd}{4}$, which is not always an overapproximation, it is possible that $imp(a)=\frac{cd}{2} > s_{FSB}(a)$, for $val(a)=0$ and $val(b)>0$.

We call {\em global scoring} (GS) these functions $s_{FSB},s_{SR}$  because they score ReLUs as accuratly as possible, considering that they do not have access to the values 
$\val(a),\val(b)$ maximizing $z$. Following this novel analysis of $s_{FSB},s_{SR}$, next Section presents a novel scoring function more accurate wrt $imp(a)$.

	}
